Comparing probabilistic predictive models applied to football

### 2015-05-08 template.tex
Alteracoes gerais e sugestoes do Rafael

Bothered by such concerns, our goal was to answer the following question: is it possible, using a simple probabilistic model, to predict results of football matches better than more sophisticated or complex models?
To answer this, we compared two sophisticated models \added{that consider several covariates} with simple multinomial-Dirichlet models that consider only the number of matches won, drawn or lost by each team as inputs.
The models were tested for the second round of the Brazilian football championships of first and second divisions and compared using standard metrics or scoring rules.
We also used other criteria such as the proportion of matches that were predicted ``correctly'' by each model and a measure of calibration, both explained in details below.
According to the adopted criteria, the simple models showed a similar predictive performance to the complex ones.

The predictions of the complex models used here as benchmarks are published on Internet websites for every matchday, and are treated by us as black-box models since we are not interested in how they were built or the assumptions on which they are based.
We know that\footnote{One of the authors is responsible for one of these models and the other was described in a dissertation.} both models are based on the bivariate Poisson or Holgate distributions, and are able to predict the final score of a match given explanatory variables.\footnote{For more details see \cite{arruda2000}.}
\deleted[remark={acho que essa parte nao eh tao simples de entender e nao eh vital para o artigo}]{Therefore, to estimate the probability that, for instance, the home team will win, the estimated model---the model estimated with available information---is simulated a given number of times and the proportion of simulated scores favoring the home team will be the predicted probability of the model.} \added{Both models output three probabilities for each match: the probability that the house team wins, the probability that 
the visitor wins, and the probability that there is a draw.} \red{usei a nomenclatura correta na frase anterior?}
\replaced{Because both}{Knowing that both} models are based on an elaborate parametric model, they are suitable for the comparison we are proposing: simple versus complex models.

There are several ways to score or classify predictions of categorical events that assume one out of three possible outcomes, like football matches.
See \cite{constantinou} for a brief survey of such measures applied to football.
We decided to score the previsions for each match in terms of their distances from the truth, i.e. the verified event, once it has occurred, and chose the most used distances in the literature: Brier (\cite{brier1950}), logarithmic and spherical.
\deleted{The total inaccuracy of each model is then defined as the sum of the point-wise scores of all its previsions.}

This paper is organized as follows.
Section \ref{sec::experimental} describes the proposed models, Section \ref{sec::scoring} presents the scoring rules and criteria used in this work to classify the models.
Section \ref{sec::results} reports the performance of the models predicting 380 matches comprising the second half of Brazilian championships of first and second divisions.
In Section \ref{sec::discussion} we discuss the results, and Section \ref{sec::conclusions} closes with some remarks and proposals for future research.
\red{talvez juntar as duas ultimas secoes?}

@@

How can we fairly distinguish better-informed states of knowledge from more poorly informed states?
\replaced{A way of doing this is to use proper scoring rules for evaluating states of uncertain knowledge.}{Using proper scoring rules for evaluating states of uncertain knowledge.}
A scoring rule is a function of two variables (or vectors, matrices etc) that \replaced{assigns}{accords} a real number (score) 
\replaced{to each
possible outcome $X$ and its respective prevision value, $P(X)$.}{
to each pair of numbers that could represent the value of $X$ and the prevision value $P(X)$.}
\replaced{ It is maximum for each $x$ when $x=P(X)$, and is nonincreasing both as $x$ departs from $P(X)$ (for each $P(X)$) and as $P(X)$ departs from $x$ (for each $x$).}{It is maximum for each $x$ when $x=P(X)$, and nonincreasing as $x$ departs from $P(X)$ for each $P(X)$, and as $P(X)$ departs from $x$ for each $x$. } \added{Thus},
the best score is achieved when knowledge of $X$ is asserted as the value of $X$ itself, whatever it might be. Most of the scoring rules \added{also} exhibit an arbitrary normalizing feature \added{such that} $S[x,P(X)]=0$ for each $x$.

Although many functions satisfy the criteria for a scoring rule specified above, not every scoring rule \replaced{can}{would} be regarded as providing a fair assessment of the assertion $P(X)$. That is, 
not all scoring rules encourage honesty and accuracy in assessing numerical values for their   previsions.
Those that do encourage a fair assessment are named \emph{proper scoring rules} \red{colocar referencia}.

\added{Say a forecaster's uncertainty about $X$ is $P_0$. A proper scoring rule is a scoring rule $S$ such that the prevision value $P(X)$ that minimizes $E_{P_0}[S(X,P(X))]$ is $P_0$. } \replaced{In other words, if one receives a reward $S(x,P(x))$ when $x$ happens and if $S$ is a proper scoring rule, then the highest expected reward is obtained by reporting one's  true uncertainty about $X$}{Imagine one receives a reward of $S(\mathbf{r},i)$ if the $i$-th event occurs.} The use of a proper scoring rule \added{thus} encourages the forecaster to be honest \deleted{to maximize the expected reward}.

The scoring rule is constructed according to the basic idea that the resulting device should oblige each participant to express his true feelings, because any departure from his own personal probability results in a diminution of his own average score as he sees it. \red{esse paragrafo me parece um pouco redundante, serah que eh melhor tirar?}

Moreover, this method leads automatically to an overall comparison between the outcomes of different personal evaluations concerned with the same totality of events.
The accumulated loss is indeed a \replaced{truly?}{throughly} concrete measure of success.

\added{In what follows, we describe in detais three proper scoring rules we use to asses the various models considered. We also 
recall the concept of calibration.}

@@

\subsection{Brier \replaced{scoring rule}{score}}

@@

The green surface \added{in Figure \ref{fig:simplex}} represents the 2-simplex, i.e. the set of points such that $p_1+p_2+p_3=1$ for non-negative values of $p_1$, $p_2$ and $p_3$.
The triangle representing the simplex has sides of length $\sqrt{2}$ and its height is $\sqrt{6}/2$.
Drawing a similar equilateral triangle with height 1 and sides $2\sqrt{3}/3$, it is possible to represent all the points of the simplex.
This new triangle is usually used to graphically display the predictions as internal points because the sum of the distances of every point inside it to each side is always one representing the probability of each event.

@@

\draw[-,thick] (-.5,0) -- (-.5,1);
\draw[-,thick] (-.45,0) -- (-.55,0);
\draw[-,thick] (-.45,1) -- (-.55,1);

@@

For every point inside this triangle, $p_1+p_2+p_3=1$.

@@

Probability assertions are said to be well calibrated at the level of probability p if the observed proportion of all propositions that are assessed with probability $p$ equals $p$.

A person's probability assertions are said to be well calibrated at the value p if the limiting frequency of occurrence of events that are assessed with probability $p$, equals $p$.

@@

We display only the results for the first division
of the Brazilian football championship; the second
division yielded similar results.
There is a total of \red{xxxx} games and \red{xxxx}
teams. \red{descrever mais alguma coisa sobre o campionato?}

@@

\hline
Matchday & BB1 & BB2 & $Mn-Dir_1$ & $Mn-Dir_2$ & $Mn-Dir_3$\\
\hline
\hline
1$^{st}$ &4.992 & 5.055 & 5.630 & 5.053 & 5.521 \\
2$^{nd}$ &5.006 & 5.328 & 5.327 & 5.533 & 5.201 \\
3$^{rd}$ &7.129 & 6.568 & 7.979 & 7.640 & 8.216 \\
4$^{th}$ &5.941 & 5.782 & 6.513 & 6.848 & 6.564 \\
5$^{th}$ &5.246 & 5.858 & 5.669 & 5.368 & 5.601 \\
6$^{th}$ &5.584 & 5.095 & 6.055 & 6.112 & 6.033 \\
7$^{th}$ &5.641 & 5.817 & 5.250 & 5.062 & 5.134 \\
8$^{th}$ &6.112 & 6.710 & 6.315 & 6.673 & 6.348 \\
9$^{th}$ &6.433 & 6.714 & 5.768 & 6.061 & 5.717 \\
10$^{th}$ &4.912 & 4.620 & 6.019 & 6.115 & 6.010 \\
\hline
Total & 111.04 & 111.32 & 115.45 & 115.36 & 115.29 \\
\hline
Mean & 0.5844 & 0.5859 & 0.6076 & 0.6072 & 0.6068 \\
\hline
\end{tabular}
\caption{Score of the ten first rounds and mean score per match after 190 matches}
\end{center}
\end{table}

@@

We evaluate how reasonable
predictions  by assessing the calibration of the methods considered,
i.e., by evaluating how often events which have assigned probability $p$ (for each $0<p<1$) happen (Section \ref{sec::scoring}).
If these probabilities are close to $p$, one concludes that the methods are well-calibrated. 
Results are displayed in Figure \ref{fig::calibration}.
Because we typically do not have several
predictions with the same assigned probability $p$, we obtain the plot by smoothing the indicator function of whether
a given event happened on  the assigned  probability of that event. The smoothing is done via smoothing splines 
\cite{wahba}, with tuning parameters chosen by cross-validation.
Because the curves obtain are close to the identity, we conclude that all methods are well-calibrate, except perhaps BB1, in which events with both low and high estimated probabilities happen more often 
than predicted.

@@

\section{Discussion}
\label{sec::discussion}

\red{talvez juntar com as conclusoes?}

@@

It is possible to interpret the score as a numerical measure of how inaccurate was the 
prevision.

\section{Conclusions}
\label{sec::conclusions}

@@

\added{This work thus indicates that while there has been a boom in the literature with the proposal of several
 fancy and involved models, especially in the era
 of big data, simple models
 can many times do as well as their complex counterparts. This is many time
 unnoticed by users of such methods, which
leads to the overuse of  models that are hard to interpret and to implement, and that also require
one to keep track of several covariates. This 
phenomenon is \red{nao consegui ligar as ideias formalmente, nao sei se vale a pena mencionar algo assim}
  ''But he isn't wearing anything at all!" Anderson's "The Emperor's New Clothes" \cite{emperor}.
Although we have explored only models for
football games,
we believe the same extends to other fields as, for instance, \red{exemplos}}

On the one hand, our findings appear as another supportive example of Ockham's razor but, on the other, \replaced{they pose}{poses} several questions about probabilistic prediction of sport events.
\replaced{In particular, based on the fact that
the models has similar performance, one may ask:}{We mention just a couple of them, raised by the similar performance of the considered models:}
 is there an irreducible ``randomness'' or ``degree of unpredictability'' implicit in these events?
Is this degree an indicator of how close is the championship being studied? \red{nao entendi bem a segunda pergunta!}
\replaced{A suggestion to future research
is to answer these questions by considering more championships and models, and  by comparing them using other scoring rules.}{To answer these questions we would have to consider more championships and models and compare them using other scoring rules, being a suggestion for further research.}
\replaced{Another direction we would like to address is to test other weighting methods 
in models 1 and 2 we propose, and to evaluate
their impact on the predictive power of the resulting predictions.}{Also as future research we would like to test other weights and study their impact on the predictive power of models 1 and 2 here proposed.}

@@

\acknowledgments{Acknowledgments}

Main text.

@@

\authorcontributions{Author Contributions}

Main text.

@@

\conflictofinterests{Conflicts of Interest}

State any potential conflicts of interest here or ``The authors declare no conflict of interest''.

@@

% Reference 1
\bibitem{ref-journal}
Lastname, F.; Author, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142-149.

% Reference 2
\bibitem{ref-book}
Lastname, F.F.; Author, T. The title of the cited contribution. In {\em The Book Title}; Editor, F., Meditor, A., Eds.; Publishing House: City, Country, 2007; pp. 32-58.

>>>>

Bothered by such concerns, our goal was to answer the following question: is it possible, using a simple probabilistic model, to predict results of football matches better than more sophisticated or complex models?
To answer this, we compared two sophisticated models that consider several covariates with simple multinomial-Dirichlet models that consider only the number of matches won, drawn or lost by each team as inputs.
The models were tested for the second round of the Brazilian football championships of first and second divisions and compared using standard metrics or scoring rules.
We also used other criteria such as the proportion of matches that were predicted ``correctly'' by each model and a measure of calibration, both explained in detail below.
According to the adopted criteria, the simple models showed a similar predictive performance to the complex ones.

The predictions of the complex models used here as benchmarks are published on Internet websites for every matchday, and are treated by us as black-box models since we are not specifically interested in how they were built or the assumptions on which they are based.
We know that both models are based on the bivariate Poisson or Holgate distributions, and are able to predict the final score of a match given explanatory variables.\footnote{One of the authors is responsible for one of these models and the other was described in a dissertation, see \cite{arruda2000}.}
Both models output three probabilities for each match: the probability that the home team wins, the probability that the away team wins, and the probability that there is a draw.
Because both models are based on an elaborate parametric assumptions, they are suitable for the comparison we are proposing: simple versus complex models.\footnote{To be precise we would have to define what is a ``simple'' model and what is a ``complex'' model. Based on the opening quotations, one may say that the number of parameters defines the complexity of a model. This is not the exactly the definition we will use since we consider other criteria, like the interpretation of explanatory variables and mathematical structure of the statistical model.}

There are several ways to score or classify predictions of categorical events that assume one out of three possible outcomes, like football matches.
See \cite{constantinou} for a brief survey of such measures applied to football.
We decided to score the previsions for each match in terms of their distances from the truth, i.e. the verified event, once it has occurred, and chose the most used distances in the literature: Brier (\cite{brier1950}), logarithmic and spherical.

This paper is organized as follows.
Section \ref{sec::experimental} describes the proposed models, Section \ref{sec::scoring} presents the scoring rules and criteria used in this work to classify the models.
Section \ref{sec::results} reports the performance of the models predicting 380 matches comprising the second half of Brazilian championships of first and second divisions.
In Section \ref{sec::remarks} we discuss the results and close with proposals for future research.

@@

How can we fairly distinguish better-informed states of knowledge from more poorly informed states?
A way of doing this is to use proper scoring rules for evaluating states of uncertain knowledge, where 
the score may be interpreted as a numerical measure of how inaccurate was the prevision.

More precisely, a scoring rule is a function of two variables (or vectors, matrices etc) that assigns a real number (score) to each possible outcome $X$ and its respective prevision value, $P(X)$.
It is maximum for each $x$ when $x=P(X)$, and is nonincreasing both as $x$ departs from $P(X)$ (for each $P(X)$) and as $P(X)$ departs from $x$ (for each $x$).
Thus, the best score is achieved when knowledge of $X$ is asserted as the value of $X$ itself, whatever it might be. Most of the scoring rules also exhibit an arbitrary normalizing feature such as $S[x,P(X)]=0$ for each $x$.

Although many functions satisfy the criteria for a scoring rule specified above, not every scoring rule can be regarded as providing a fair assessment of the assertion $P(X)$, that is, not all scoring rules encourage honesty and accuracy in assessing numerical values for their previsions.
Those that do encourage a fair assessment are named \emph{proper scoring rules} \red{colocar referencia}.

Say a forecaster's uncertainty about $X$ is $P_0$. A proper scoring rule is a scoring rule $S$ such that the prevision value $P(X)$ that minimizes $E_{P_0}[S(X,P(X))]$ is $P_0$. 
In other words, if one receives a reward $S(x,P(x))$ when $x$ happens and if $S$ is a proper scoring rule, then the highest expected reward is obtained by reporting one's  true uncertainty about $X$.
The use of a proper scoring rule thus encourages the forecaster to be honest to maximize the expected reward.
Moreover, this method leads automatically to an overall comparison between the outcomes of different personal evaluations concerned with the same totality of events.
Therefore, the accumulated score may be considered a concrete measure of success of each forecaster or model.

In what follows, we describe in detail three proper scoring rules we use to asses the various models considered. 
We also recall the concept of calibration.

@@

\subsection{Brier scoring rule}

@@

The green surface in Figure \ref{fig:simplex} represents the 2-simplex, i.e. the set of points such that $p_1+p_2+p_3=1$ for non-negative values of $p_1$, $p_2$ and $p_3$.
The triangle representing the simplex has sides of length $\sqrt{2}$ and its height is $\sqrt{6}/2$.
Drawing a similar equilateral triangle with height 1 and sides $2\sqrt{3}/3$, it is possible to represent all the points of the simplex.
This new triangle is used to graphically display the predictions as internal points because the sum of the distances of every point inside it to each side, representing the probability of each event, is always one.

@@

\draw[-,thick] (-.5,0) -- (-.5,1);
\draw[-,thick] (-.45,0) -- (-.55,0);
\draw[-,thick] (-.45,1) -- (-.55,1);
\draw[dashed,thick] (-.5,1) -- (0.57735,1);
\draw[dashed,thick] (-.5,0) -- (0,0);

@@

@@

Probability assertions are said to be well calibrated at the level of probability $p$ if the observed proportion of all propositions that are assessed with probability $p$ equals $p$.

A person's probability assertions are said to be well calibrated at the value $p$ if the limiting frequency of occurrence of events that are assessed with probability $p$, equals $p$.

@@

We display only the results for the first division of the Brazilian football championship; the second
division yielded similar results.
Both divisions have 20 teams that play against each other twice (home and away) and the team with more points after all matches were played is declared champion.
The last four are relegated to a minor division and, in first division, the first four play Copa Libertadores (South America champions league) and in second division the first four are promoted to first division.
Therefore, in each championship there is a total of 380 matches, 190 in each round.

@@

\hline
Matchday & BB1 & BB2 & $Mn-Dir_1$ & $Mn-Dir_2$ & $Mn-Dir_3$\\
\hline
\hline
1 &4.992 & 5.055 & 5.630 & 5.053 & 5.521 \\
2 &5.006 & 5.328 & 5.327 & 5.533 & 5.201 \\
3 &7.129 & 6.568 & 7.979 & 7.640 & 8.216 \\
4 &5.941 & 5.782 & 6.513 & 6.848 & 6.564 \\
5 &5.246 & 5.858 & 5.669 & 5.368 & 5.601 \\
\hline
Total & 111.04 & 111.32 & 115.45 & 115.36 & 115.29 \\
\hline
Mean & 0.5844 & 0.5859 & 0.6076 & 0.6072 & 0.6068 \\
\hline
\end{tabular}
\caption{Score of the five first matchdays and mean score per match after 190 matches}
\end{center}
\end{table}

@@

We evaluate how reasonable \added{were the?} predictions by assessing the calibration of the methods considered, i.e., by evaluating how often events which have assigned probability $p$ (for each $0<p<1$) happen \added{or happened?} (Section \ref{sec::scoring}).
If these probabilities are close to $p$, one concludes that the methods are well-calibrated. 
Results are displayed in Figure \ref{fig::calibration}.
Because we typically do not have several predictions with the same assigned probability $p$, we obtain the plot by smoothing the indicator function of whether a given event happened on  the assigned  probability of that event. 
The smoothing is done via smoothing splines \cite{wahba}, with tuning parameters chosen by cross-validation.
Because the curves obtain \added{(ou we obtained?)} are close to the identity, we conclude that all methods are well-calibrated, except perhaps BB1, in which events with both low and high estimated probabilities happen more often 
than predicted.

@@

\section{Final remarks}
\label{sec::remarks}

@@

@@

This work thus indicates that while there has been a boom in the literature with the proposal of several fancy and involved models, especially in the era of big data, simple models can many times do as well as their complex counterparts. 
This is often unnoticed by users of such methods, which leads to the overuse of models that are hard to interpret and to implement, and that also require one to keep track of several covariates.
We would call this phenomenon the {\it king's nudity}\footnote{In allusion to Anderson's {\it The Emperor's New Clothes}, \cite{emperor}: ``But he isn't wearing anything at all!''.}, that is not, of course, always observed.
In this spirit, since we have explored only the prediction of football matches, we would like to study if the the same extends to other fields such as, for instance, \red{exemplos} \red{Tentei aqui Rafael, veja o que acha.}

If, on the one hand, our findings appear as another supportive example of Ockham's razor, on the other they pose several questions about probabilistic prediction of sport events.
In particular, based on the fact that the models have similar performance, one may ask: is there an irreducible ``randomness'' or ``degree of unpredictability'' implicit in these events?
Is this degree an indicator of how tight or leveled is the championship being studied? \red{ficou mais clara a segunda pergunta? Queria perguntar se a dificuldade de prever os resultados indica o quao disputado ou equilibrado \'e o campeonato.}

A suggestion of future research is to answer these questions by considering more championships and models, and  by comparing them using other scoring rules.
Another direction we would like to address is to test other weighting methods in models 1 and 2 here proposed, and to evaluate their impact on the predictive power of the resulting predictions.

@@

%\acknowledgments{Acknowledgments}

%Main text.

@@

%\authorcontributions{Author Contributions}

%Main text.

@@

%\conflictofinterests{Conflicts of Interest}

%State any potential conflicts of interest here or ``The authors declare no conflict of interest''.

@@

% Reference 1
%\bibitem{ref-journal}
%Lastname, F.; Author, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142-149.

% Reference 2
%\bibitem{ref-book}
%Lastname, F.F.; Author, T. The title of the cited contribution. In {\em The Book Title}; Editor, F., Meditor, A., Eds.; Publishing House: City, Country, 2007; pp. 32-58.
